---
title: Question 4
author: "Sarvesh Bhagavathi, 33716374"
format: closeread-html
---

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(
  echo = FALSE,       
  warning = FALSE,   
  message = FALSE,   
  fig.width = 12,     
  fig.height = 10     
)
```

## When the Roof Caved In: Lessons from the 2007–2008 Housing Crisis

Governments are tearing down public housing, placing blind faith in the idea that the market will “fix everything.” But what happens when the market itself collapses? The 2007–2008 financial crisis showed us exactly that. Housing prices plummeted, communities suffered, and the very market forces they trusted became the root of the disaster. If we keep relying on market-driven solutions without safeguarding affordable housing, are we just setting ourselves up for another catastrophe? Here’s a classic example of what happens when policies fail to prioritize people over profits.

::: cr-section
This line graph depicts the average house prices during the financial crisis of 2007–2008. In early 2007, house prices were stable, but by mid-2007, the trend sharply reversed. @cr-plot1 The dashed red line marks the onset of the crisis, highlighting a steep decline in prices—a drop of over \$300,000 in less than two years.

This crisis wasn’t just a blip; it was a systemic failure driven by unchecked financial speculation and housing market overreach. The result? A devastating blow to millions of homeowners.

::: {#cr-plot1}
```{r}

library(tidyverse)
library(scales)
library(lubridate)
library(zoo)

house_sales <- read_csv(here::here("data/house-sales.csv"))
```

```{r}
zoomed_data <- house_sales %>%
  filter(date >= as.Date("2007-01-01") & date <= as.Date("2008-11-16")) %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarize(avg_price = mean(price, na.rm = TRUE))

ggplot(zoomed_data, aes(x = month, y = avg_price)) +
  geom_line(color = "blue", size = 1.2) +
  geom_vline(xintercept = as.Date("2007-09-01"), linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = as.Date("2007-11-29"), y = 980000, 
           label = "Crisis Hits the Housing Market", 
           color = "red", size = 4, hjust = 0.5, fontface = "bold") +
  labs(
    title = "House Prices During the Financial Crisis (2007–2008)",
    x = "Year",
    y = "Average Price"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(size = 10, hjust = 0)
  )
```
:::

This graph zooms into Alameda and Solano counties to showcase their similar yet slightly contrasting experiences during the housing crisis. Alameda, with its higher median home prices, showed resilience better recovery compared to Solano. @cr-plot2

The shaded red region marks the crisis peak—a period when both counties experienced the harshest depreciation in value.

::: {#cr-plot2}
```{r}

selected_data <- house_sales %>%
  filter(
    date >= as.Date("2007-01-01") & date <= as.Date("2008-11-16"),
    county %in% c("Alameda County", "Solano County")
  ) %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(county, month) %>%
  summarize(avg_price = mean(price, na.rm = TRUE)) %>%
  arrange(county, month)

selected_data <- selected_data %>%
  group_by(county) %>%
  mutate(smoothed_price = rollmean(avg_price, k = 3, fill = NA, align = "right"))

# Plotting
ggplot(selected_data, aes(x = month, y = smoothed_price, color = county)) +
  geom_line(size = 1.2) +
  annotate("rect", xmin = as.Date("2007-10-01"), xmax = as.Date("2008-05-01"),
           ymin = -Inf, ymax = Inf, fill = "lightpink", alpha = 0.3) +
  annotate("text", x = as.Date("2008-01-01"), y = 800000,
           label = "Crisis Peak", color = "blue", size = 4, fontface = "bold") +
  labs(
    title = "Smoothed House Prices for Alameda and Solano Counties (2007–2008)",
    x = "Month",
    y = "Smoothed Average Price (3-Month Moving Average)",
    color = "County"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "bottom",
    plot.caption = element_text(size = 13, hjust = 0, face = "italic"),
    plot.margin = margin(t = 10, r = 10, b = 30, l = 10), # Add more space for caption
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  coord_cartesian(clip = "off") # Ensures space for large caption

```
:::

This graph tells a deeper story of privilege: the wealthy were shielded from the brunt of the crisis, while less affluent regions bore the heaviest losses. @cr-plot3

Counties with more affordable homes, like Napa, experienced the steepest declines in property prices, with a dramatic 36.7% drop.

Rather than addressing these structural inequities, the market’s response further entrenched them, rewarding those who could afford to wait out the storm while devastating those who couldn’t.

This discrepancy highlights the need for proactive housing policies that address regional vulnerabilities. Strengthening affordable housing programs can shield lower-income markets during crises.

::: {#cr-plot3}
```{r}

house_sales <- house_sales %>%
  mutate(
    Period = case_when(
      date >= as.Date("2005-01-01") & date < as.Date("2007-01-01") ~ "Pre-Crisis",
      date >= as.Date("2007-01-01") & date <= as.Date("2008-11-16") ~ "Crisis"
    )
  )

# Filter for valid counties
valid_counties <- house_sales %>%
  group_by(county, Period) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Period, values_from = count, values_fill = 0) %>%
  filter(`Pre-Crisis` > 0 & Crisis > 0) %>%
  pull(county)

house_sales_filtered <- house_sales %>%
  filter(county %in% valid_counties & !is.na(Period))

# Calculate percentage decline for annotations
median_prices <- house_sales_filtered %>%
  group_by(county, Period) %>%
  summarize(median_price = median(price, na.rm = TRUE), .groups = "drop")

median_pre_crisis <- median_prices %>% filter(Period == "Pre-Crisis")
median_crisis <- median_prices %>% filter(Period == "Crisis")

median_combined <- median_pre_crisis %>%
  inner_join(median_crisis, by = "county", suffix = c("_pre", "_crisis")) %>%
  mutate(percentage_decline = (median_price_pre - median_price_crisis) / median_price_pre * 100)

largest_decline <- median_combined %>% slice_max(percentage_decline, n = 1)
smallest_decline <- median_combined %>% slice_min(percentage_decline, n = 1)

# Highlighting logic for Napa County and Marin County
highlight_data <- house_sales_filtered %>%
  filter(county %in% c("Napa County", "Marin County"))

# Plot
ggplot(house_sales_filtered, aes(x = county, y = price, fill = Period)) +
  geom_boxplot(outlier.shape = NA) +  # Base boxplot for all counties
  geom_boxplot(
    data = highlight_data,  # Highlighted boxplot overlay
    aes(x = county, y = price, fill = Period),
    color = "black", size = 1.2, outlier.shape = NA
  ) +
  annotate(
    "text",
    x = match(largest_decline$county, unique(house_sales_filtered$county)) - 0.2, # Adjust left
    y = 5,
    label = paste0(largest_decline$county, "\nLargest Decline: ", round(largest_decline$percentage_decline, 2), "%"),
    color = "red",
    size = 3,
    angle = 45,
    hjust = 0,
    fontface = "bold" # Add bold styling
  ) +
  annotate(
    "text",
    x = match(smallest_decline$county, unique(house_sales_filtered$county)) - 0.2, # Adjust left
    y = 5,
    label = paste0(smallest_decline$county, "\nSmallest Decline: ", round(smallest_decline$percentage_decline, 2), "%"),
    color = "blue",
    size = 3,
    angle = 45,
    hjust = 0,
    fontface = "bold" # Add bold styling
  ) +
  labs(
    title = "Box Plot of House Prices by County (Pre-Crisis vs. Crisis)",
    x = "County",
    y = "Price (USD)"
  ) +
  scale_y_continuous(
    limits = c(0, 2000000),
    breaks = seq(0, 2000000, by = 200000),
    labels = comma
  ) +
  scale_fill_manual(
    values = c("Pre-Crisis" = "blue", "Crisis" = "red"),
    name = "Period"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

```
:::

This density plot compares house price distributions before and during the crisis. The overlap highlights the few stable price ranges, while the crisis distribution shifts leftward, indicating plummeting values. @cr-plot4

The steep decline in prices pushed many homes into lower price brackets, with far-reaching effects on homeowners’ equity and financial security.

::: {#cr-plot4}
```{r}

mean_prices <- house_sales_filtered %>%
  group_by(Period) %>%
  summarize(mean_price = mean(price, na.rm = TRUE))

# Extract means for Pre-Crisis and Crisis
mean_pre_crisis <- mean_prices %>% filter(Period == "Pre-Crisis") %>% pull(mean_price)
mean_crisis <- mean_prices %>% filter(Period == "Crisis") %>% pull(mean_price)

# Plot with means
ggplot(house_sales_filtered, aes(x = price, fill = Period)) +
  geom_density(alpha = 0.5) +  # Semi-transparent to show overlap
  geom_vline(xintercept = mean_pre_crisis, color = "blue", linetype = "dashed", size = 1.2) + 
  geom_vline(xintercept = mean_crisis, color = "red", linetype = "dashed", size = 1.2) +
  annotate(
    "text",
    x = mean_pre_crisis, y = 1.5e-06,
    label = "Pre-Crisis Mean",
    color = "blue",
    size = 4,
    angle = 90,
    vjust = -0.5,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = mean_crisis, y = 1.5e-06,
    label = "Crisis Mean",
    color = "red",
    size = 4,
    angle = 90,
    vjust = -0.5,
    fontface = "bold"
  ) +
  labs(
    title = "Distribution of House Prices (Pre-Crisis vs. Crisis)",
    x = "Price (USD)",
    y = "Density"
  ) +
  scale_x_continuous(labels = comma, limits = c(0, 2000000)) +
  scale_fill_manual(
    values = c("Pre-Crisis" = "blue", "Crisis" = "red"),
    name = "Period"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

```
:::

This visualization emphasizes more on how middle and lower price brackets absorbed the brunt of the housing market’s turmoil. @cr-plot5

Here, the purple-shaded region highlights the overlapping price ranges that remained stable during the crisis. The non-overlapping tails reflect the dramatic shifts in affordability.

Isn’t it worth to ask ourseves why should access to a basic human need like housing be tied to a speculative market that fails at times - the very people it’s supposed to serve?”

::: {#cr-plot5}
```{r}
density_pre <- density(house_sales_filtered$price[house_sales_filtered$Period == "Pre-Crisis"], from = 0, to = 2000000)
density_crisis <- density(house_sales_filtered$price[house_sales_filtered$Period == "Crisis"], from = 0, to = 2000000)

# Create a data frame with the density values and compute the overlap
overlap_data <- data.frame(
  x = density_pre$x,
  y_pre = density_pre$y,
  y_crisis = density_crisis$y
)
overlap_data$y_overlap <- pmin(overlap_data$y_pre, overlap_data$y_crisis)

# Plot with the shaded overlapping region
ggplot() +
  # Pre-Crisis density
  geom_line(data = overlap_data, aes(x = x, y = y_pre), color = "blue", size = 1) +
  # Crisis density
  geom_line(data = overlap_data, aes(x = x, y = y_crisis), color = "red", size = 1) +
  # Shaded overlap
  geom_ribbon(data = overlap_data, aes(x = x, ymin = 0, ymax = y_overlap), fill = "purple", alpha = 0.3) +
  labs(
    title = "Distribution of House Prices (Pre-Crisis vs. Crisis)",
    x = "Price (USD)",
    y = "Density"
  ) +
  scale_x_continuous(labels = scales::comma, limits = c(0, 2000000)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(size = 10)
  )

```
:::

The increase in affordable homes wasn’t a victory—it was a symptom of systemic collapse. This wasn’t about improving accessibility but about economic distress forcing values to plummet. @cr-plot6

This bar chart illustrates how sales shifted during the crisis. Pre-crisis, mid-range homes dominated the market. During the crisis, homes under \$500k accounted for a much larger share of sales. This shift reflects the mass devaluation of homes, as prices fell, pushing many properties disguised ine the name of the “affordable range.”

::: {#cr-plot6}
```{r}


house_sales <- house_sales %>%
  mutate(
    Period = case_when(
      date >= as.Date("2005-01-01") & date < as.Date("2007-01-01") ~ "Pre-Crisis",
      date >= as.Date("2007-01-01") & date <= as.Date("2008-11-16") ~ "Crisis"
    )
  )

house_sales_filtered <- house_sales %>%
  filter(!is.na(Period))

house_sales_filtered <- house_sales_filtered %>%
  mutate(
    Price_Range = case_when(
      price < 500000 ~ "<500k",
      price >= 500000 & price <= 1000000 ~ "500k–1M",
      price > 1000000 ~ ">1M"
    )
  )

proportion_data <- house_sales_filtered %>%
  group_by(Period, Price_Range) %>%
  summarize(Count = n()) %>%
  group_by(Period) %>%
  mutate(Proportion = Count / sum(Count))

ggplot(proportion_data, aes(x = Period, y = Proportion, fill = Price_Range)) +
  geom_bar(stat = "identity", position = "stack", color = "black") +
  labs(
    title = "Proportion of Houses Sold by Price Range (Pre-Crisis vs. Crisis)",
    x = "Period",
    y = "Proportion of Sales",
    fill = "Price Range"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

```
:::

<!-- This ends the scrollytelling section -->
:::
